Page not found

### 2023-06-21 manuscript/appendices/B_mean_patterns.qmd
goodbye i will now sleep for 3 days

[^details]: We are glossing over the fact that $S^n : [0,1]^2 \mapsto [0,1]$, but the positions $\vec{z_i} \in \R^2$. In the code, we can deal with these by either applying the inverse CDF (as described in @sec-mapping-cdf), or considering a bounding box $[\min_i(z_{1,i}), \max_i(z_{1,i})] \times [\min_i(z_{2,i}), \max_i(z_{2,i})]$ of the sampled points, which is then rescaled to $[0,1] \times [0,1]$ (see implementation for [`Box`](https://github.com/ninivert/lcnthesis/blob/master/notebooks/neurodyn/_mapping.py#L24)). Both approaches are equivalent, in the sense that they give rise to the same dynamics as $n$ becomes large, although the obtained connectivity matrix $\tilde J_{\alpha\beta}$ may not be the identical.

@@

* @fig-kernels-1 is the method described in this appendix, based on sampling $N$ neurons in the PDF space $\R^2$, applying the mapping on the obtained positions $\vec{z_i}$, and finally applying the binning (this is coarse-graining). We see, through the empty rows in the connectivity, that due to finite-$N$ effects, locations where $\rho$ is small are not represented, and, as a consequence, no neurons are present in the corresponding 1D bin. When this happens, we set these rows to zero.
* To obtain @fig-kernels-2, we make the approximation that the support of $\rho(z_1, z_2)$ is compact in $[-4,4]\times[-4,4]$, which of course is not true, but in the case of a Gaussian distribution this is a reasonable approximation, because $\CDF(4) \approx 0.999968$. Defining $\tilde \rho = \rho \circ S^{-1}$, the grid is numerically renormalized so that despite the compact support approximation, the probabilities sum up to one, and the connectivity matrix is modified so that the columns are weighed by $\tilde \rho(\beta)$:
  $$
  \tilde J_{\alpha\beta} = \tilde \rho(\beta) \sum_{\mu=1}^p \tilde F_{\mu \alpha} \tilde G_{\mu \beta}
  $$
  We see that compared to @fig-kernels-1, using the grid method allows us to "sample" the locations where the density is small, and this results in less "whited-out" *rows*. Still, vertical bands of very fade colour correspond to the reweighing of the *columns* where $\tilde \rho(\beta)$ is small.
* Finally, @fig-kernels-3 shows the kernel obtained from applying the grid method in the CDF space $[0,1]^2$, as described in @sec-mapping-cdf. We see that the connectivity seems much more "uniform", because each point of the grid has "equal weight" in the CDF space.

>>>>

[^details]: We are glossing over the fact that $S^n : [0,1]^2 \mapsto [0,1]$, but the positions $\vec{z_i} \in \R^2$. In the code, we can deal with this by either applying the inverse CDF (as described in @sec-mapping-cdf), or considering a bounding box $[\min_i(z_{1,i}), \max_i(z_{1,i})] \times [\min_i(z_{2,i}), \max_i(z_{2,i})]$ of the sampled points, which is then rescaled to $[0,1] \times [0,1]$ (see implementation for [`Box`](https://github.com/ninivert/lcnthesis/blob/master/notebooks/neurodyn/_mapping.py#L24)). Both approaches are equivalent, in the sense that they give rise to the same dynamics as $n$ becomes large, although the obtained connectivity matrix $\tilde J_{\alpha\beta}$ may not be the identical.

@@

* @fig-kernels-1 is the method described in this appendix, based on sampling $N$ neurons in the PDF space $\R^2$, applying the mapping on the obtained positions $\vec{z_i}$, and finally applying the binning (this is coarse-graining). We see, through the empty rows in the connectivity, that due to finite-$N$ effects, locations where $\rho$ is small are not represented, and, as a consequence, no neurons are present in the corresponding 1D bin. When this happens, we set these rows to zero.
* To obtain @fig-kernels-2, we make the approximation that the support of $\rho(z_1, z_2)$ is compact in $[-4,4]\times[-4,4]$, which of course is not true, but in the case of a Gaussian distribution this is a reasonable approximation, because $\CDF(4) \approx 0.999968$. Defining $\tilde \rho = \rho \circ S^{-1}$, the grid is numerically renormalized so that despite the compact support approximation, the probabilities sum up to one, and the connectivity matrix is modified so that the columns are weighed by $\tilde \rho(\beta)$:
  $$
  \tilde J_{\alpha\beta} = \tilde \rho(\beta) \sum_{\mu=1}^p \tilde F_{\mu \alpha} \tilde G_{\mu \beta}.
  $$
  We see that compared to @fig-kernels-1, using the grid method allows us to "sample" the locations where the density is small, and this results in less "whited-out" *rows*. Still, vertical bands of very fade colour correspond to the reweighing of the *columns* where $\tilde \rho(\beta)$ is small.
* Finally, @fig-kernels-3 shows the kernel obtained from applying the grid method in the CDF space $[0,1]^2$, as described in @sec-mapping-cdf. We see that the connectivity seems much more "uniform", because each point of the grid has "equal weight" in the CDF space.

### 2023-06-21 manuscript/chapters/1_intro.qmd
finished figures !!! finally
goodbye i will now sleep for 3 days

Early recordings of cat's somatosensory @Mou57 and visual @HubWie62 cortices suggested that the cortical sheet is organized in vertical columns of functionally similar neurons.

The putative columnar organization of the cortex motivated the design of spatially structured models of neural population dynamics, namely neural field models [@WilCow73;@Nun74;@Ama77], where the spatial dimensions corresponded to the two dimensions of the cortical sheet.

@@

More recent recording methods, especially in rodents, have provided evidence of low-dimensional organization of neuronal activity that does not depend on the physical locations of neurons on the cortical sheet. For example, contrary to the cat's visual cortex, the orientation selectivity of pyramidal neurons in a rat's visual cortex does not depend on their locations, that is, orientation selectivity is heterogeneous at any given location @OhkChu05. Moreover, the activity of interneurons in any small volume of the visual cortex is also heterogeneous and seems to be structured by a low-dimensional transcriptomic manifold of fine cell subtypes @BugDuf22. The local functional heterogeneity of neuronal activity in the cortex challenges the old concept of functionally homogeneous cortical columns and, with it, classical neural field models.

@@

The goal of this Master Thesis is to show that the answer to the second question is highly nontrivial because even the dimensionality of the embedding space, for some given spatiotemporal dynamics, is not clearly defined. This means that it is possible to find two embeddings (and connectivity kernels) of different dimensions that give rise to identical dynamics.
<!-- ; *contradicting the intuition that the embedding dimension should at least be as high as the dimensionality of the dynamics* (the number of variables needed to fully describe the dynamical system). -->

@@

![Every neuron in a rate neural network has a position in the $p$-dimensional embedding space](figures/fig-rnn.svg){width=500px}

@@

This work was inspired by recent mathematical development in mean-field theories involving the theory of graphons @JabPoy21.
Informally, these recent results suggest that, even in the absence of any explicit spatial structure, as long as weights scale as $1/N$, the dynamics of large networks can be described, as $N \to \infty$, by a neural field equation where the embedding is simply the interval $[0,1]$, which motivates the search for a formulation of a one-dimensional neural embedding.
<!-- Their main result is that under some boundedness assumptions of the connectivity between the components of a system, the equation of evolution incorporates only an integral over $[0,1]$. This is indeed the case in the toy model described next chapter, where weights scale as $J_{ij} \sim \frac 1N$, which motivates the search for a formulation of a one-dimensional neural embedding. -->

@@

2. A discretization-based simulation. In the $\Rp$ space, we discretize each dimension, forming a grid of $p$-dimensional bins. Each bin effectively corresponds to an infinite population of neurons, which are approximated to be the same (this is a mean-field approximation). This works because by continuity of the field, neurons which are close in $\Rp$ also have similar potentials. As we increase the number of bins, the simulation becomes more precise and the dynamics approach that of the true mathematical neural field.

With this in mind, we can try to understand why simulating a neural field in $[0,1]$ might be difficult. Suppose we take bins in the $\Rp$ space, and map each bin to a random position in $[0,1]$. Intuitively, this won't work (in the sense that the dynamics in $[0,1]$ will not be the same as the dynamics in $\Rp$), because the hypothesis of continuity is not longer true: the mapping is random, so the nearby neurons in $[0,1]$ will no longer have similar potentials, and the error of the mean-field approximation will no longer vanish as we take increasingly many (and smaller) bins.
<!-- NOTE : was "populations" before, this is explained later in the thesis. but OK, for the introduction, maybe we should use neurons
the mapping is random, so the nearby populations in $[0,1]$ -->

![(TODO) Good mappings from $\Rp$ to $[0,1]$ should maintain some sense of "locality"](figures/photo_2023-06-13_18-15-50.jpg){width=500px}

>>>>

Early recordings of cats' somatosensory @Mou57 and visual @HubWie62 cortices suggested that the cortical sheet is organized in vertical columns of functionally similar neurons.

The columnar organization of the cortex motivated the design of spatially structured models of neural population dynamics, namely neural field models [@WilCow73;@Nun74;@Ama77], where the spatial dimensions corresponded to the two dimensions of the cortical sheet.

@@

More recent recording methods, especially in rodents, have provided evidence of low-dimensional organization of neuronal activity that does not depend on the physical locations of neurons on the cortical sheet. For example, contrary to cats' visual cortex, the orientation selectivity of pyramidal neurons in a rat's visual cortex does not depend on their locations, that is, orientation selectivity is heterogeneous at any given location @OhkChu05. Moreover, the activity of interneurons in any small volume of the visual cortex is also heterogeneous and seems to be structured by a low-dimensional manifold of fine cell subtypes @BugDuf22. The local functional heterogeneity of neuronal activity in the cortex challenges the old concept of functionally homogeneous cortical columns and, with it, classical neural field models.

@@

The goal of this Master Thesis is to show that the answer to the second question is highly nontrivial because even the dimensionality of the embedding space, for some given spatiotemporal dynamics, is not clearly defined. This means that it is possible to find two embeddings (each associated with a connectivity kernel) of different dimensions that give rise to identical dynamics.
<!-- ; *contradicting the intuition that the embedding dimension should at least be as high as the dimensionality of the dynamics* (the number of variables needed to fully describe the dynamical system). -->

@@

![Every neuron in the neural network has a position in the $p$-dimensional embedding space.](figures/fig-rnn.svg){width=500px}

@@

This work was inspired by recent mathematical development in mean-field theories involving the theory of graphons @JabPoy21.
Informally, these recent results suggest that, as long as weights scale as $1/N$, the dynamics of large networks can be described, as $N \to \infty$, by a neural field equation where the embedding is simply the interval $[0,1]$, which motivates the search for a formulation of a one-dimensional neural embedding.
<!-- Their main result is that under some boundedness assumptions of the connectivity between the components of a system, the equation of evolution incorporates only an integral over $[0,1]$. This is indeed the case in the toy model described next chapter, where weights scale as $J_{ij} \sim \frac 1N$, which motivates the search for a formulation of a one-dimensional neural embedding. -->

@@

2. A grid-based simulation. In the $\Rp$ space, we discretize each dimension, forming a grid of $p$-dimensional bins. Each bin effectively corresponds to an infinite population of neurons, which are approximated to be the same (this is a mean-field approximation). This works because by continuity of the field, neurons which are close in $\Rp$ also have similar potentials. As we increase the number of bins, the simulation becomes more precise and the dynamics approach that of the true mathematical neural field.

With this in mind, we can try to understand why simulating a neural field in $[0,1]$ might be difficult. Suppose we take bins in the $\Rp$ space, and map each bin to a random position in $[0,1]$. Intuitively, this won't work (in the sense that the dynamics in $[0,1]$ will not be the same as the dynamics in $\Rp$), because the hypothesis of continuity is not longer true: the mapping is random, so the nearby neurons in $[0,1]$ will no longer have similar potentials, and the error of the mean-field approximation will no longer vanish as we take increasingly many (and smaller) bins. This is visualized in @fig-intuition-locality.
<!-- NOTE : was "populations" before, this is explained later in the thesis. but OK, for the introduction, maybe we should use neurons
the mapping is random, so the nearby populations in $[0,1]$ -->

![Good mappings from $\Rp$ to $[0,1]$ should maintain some sense of "locality".](figures/fig-locality.svg){width=600px #fig-intuition-locality}

### 2023-06-21 manuscript/chapters/2_neural_fields.qmd
goodbye i will now sleep for 3 days

The toy model we introduce is an instance of a low-rank neural network. Such models hold their name from the form of their connectivity matrix, which has a rank of (at most) $p$. We write the connectivity matrix as

@@

The geometric view of these networks is that the recurrent currents lie in a $p$-dimensional subspace of $\R^N$ spanned by the vectors $\{\vec{F_1}, \cdots, \vec{F_p}\}$ (where $\vec{F_\mu}=(F_{\mu 1}, \cdots F_{\mu N})$), which therefore define a "natural" linear embedding (in the sense of @JazOst21) of the neural population activity.

@@

Instead of numerically computing the full connectivity matrix $J$ (which would require storing $N^2 = 50,000^2 \sim 10^{10}$ entries, taking about 40 GiB if we use 32-bit floating point numbers... yikes !), we take advantage of its low-rank structure to compute recurrent currents, therefore only storing $p \times N$ numbers instead of $N^2$.[^code2]

@@

We however note that in this case, there are two pattern fixed points that the network can converge to. In the example given here, the network coincidentally converged again to the first pattern fixed point. The network is in the $\mu=1$ pattern, which we can see because the neural field is constant in the $z_2$ direction. We write the fixed point $h^\star(z_1, z_2, t) = z_1$. The neural field is now written as

@@

Due to the formulation of the neural field weighed by a probability distribution $\rho(y_1, \cdots, y_p)$, a natural way to estimate the integral is by Monte Carlo integration.
The essence of this method is to use the Central Limit Theorem, where we take $N$ independent samples $\vec{y_i} = (y_{1i}, \cdots, y_{pi})$ from the distribution $\rho$, and use them to estimate the integrand.

@@

Overlap variables in continuous time variables were introduced by @Gerstner_1992 to study spiking networks models of associative memory (Hopfield-type networks), which can be seen as a low-rank network with discrete spatial structure instead of our continuous $\vec{z}$. The overlap variables measure the correlation between the state of the network (the value of the potential $h(\vec z, t)$ at each point in space) and the pattern fixed point.

@@

$$
\begin{aligned}
m_\mu(t) &= \frac 1N \sum_{i=1}^N \tilde \phi(z_{\mu i}) \phi(h_i(t)) \\
&\xrightarrow{N \to \infty} \int_{\Rp} \tilde \phi(z_\mu) \phi(h(\vec z, t)) \mathcal{N}^p(\d \vec z)
\end{aligned}
$$ {#eq-overlap}

@@

$$
h(\vec z, t) = h^\perp(\vec z, t) + \sum_{\mu=1}^p \kappa_\mu(t) z_\mu
$$

>>>>

The toy model we introduce is an instance of a low-rank neural network. Such models hold their name from the form of their connectivity matrix, which has a rank of (at most) $p$. We write the connectivity matrix as follows:

@@

The geometric view of these networks is that the recurrent currents lie in a $p$-dimensional subspace of $\R^N$ spanned by the vectors $\{\vec{F_1}, \cdots, \vec{F_\mu},\cdots, \vec{F_p}\}$ (where each $\vec{F_\mu}=(F_{\mu 1}, \cdots F_{\mu N})$), which therefore define a "natural" linear embedding (in the sense of @JazOst21) of the neural population activity.

@@

Instead of numerically computing the full connectivity matrix $J$ (which would require storing $N^2 = 50,000^2 \sim 10^{10}$ entries, taking about 40 GiB if we use 32-bit floating point numbers... yikes!), we take advantage of its low-rank structure to compute recurrent currents, therefore only storing $p \times N$ numbers instead of $N^2$.[^code2]

@@

We however note that in this case, there are two pattern fixed points that the network can converge to. In the example given here, the network coincidentally converged again to the first pattern fixed point. We see this because the neural field is constant in the $z_2$ direction. We write the fixed point $h^\star(z_1, z_2, t) = z_1$. The neural field is now written as

@@

Due to the formulation of the neural field weighed by a probability distribution $\rho(y_1, \cdots, y_p)$, a natural way to estimate the integral is by Monte Carlo integration.
The essence of this method is to use the Central Limit Theorem, where we take $N$ independent samples $\vec{y_i} = (y_{1i}, \cdots, y_{pi})$ from the distribution $\rho$, and use them to estimate the integrand $I(\vec z, \vec y)$.

@@

Overlap variables in continuous time were introduced by @Gerstner_1992 to study spiking networks models of associative memory (Hopfield-type networks), which can be seen as a low-rank network with discrete spatial structure instead of our continuous $\vec{z}$. The overlap variables measure the correlation between the state of the network (the value of the potential $h(\vec z, t)$ at each point in space) and the pattern fixed point.

@@

$$
\begin{aligned}
m_\mu(t) &= \frac 1N \sum_{i=1}^N \tilde \phi(z_{\mu i}) \phi(h_i(t)) \\
&\xrightarrow{N \to \infty} \int_{\Rp} \tilde \phi(z_\mu) \phi(h(\vec z, t)) \mathcal{N}^p(\d \vec z).
\end{aligned}
$$ {#eq-overlap}

@@

$$
h(\vec z, t) = h^\perp(\vec z, t) + \sum_{\mu=1}^p \kappa_\mu(t) z_\mu.
$$
